Red-Giant Hot Jupiters: Brilliant Radio Beacons?

### 2015-09-02 ms.tex
in Section 4.3, the order of descriptions for SKA and VLA are switched in order to be consistent with the Table
updates including comments from Yuka on the sensitivity section

%\memoYF{discussion about detectability... Figures 1-3 of \citet{griesmeier2007b} show $10^{-3}$ Jy limit for 10-100 MHz (LWA, LOFAR) and $10^{-5}-10^{-4}$ Jy for 100-1000 MHz (LOFAR, SKA).}

We consider in this section several current and near future radio observatories that 
operate at frequencies $\nu \lesssim 350~\rm MHz$ and can reach continuum 
sensitivities less than 1~mJy in a few hours.  The results are summarized
in Table~\ref{tab:sens}.
The primary limitations to detectability are noise, characterized by the
system equivalent blackbody temperature $T_{\rm sys}$, resolution, and source confusion.  
To estimate the source confusion $\sigma_c$ (in mJy beam$^{-1}$) reported in Table~\ref{tab:sens}, 
we use the relations provided by \cite{condon+ransom2016}, reproduced below:
\begin{eqnarray}
\sigma_c & \approx & 0.2 \left( \frac{\nu}{1 \rm~GHz} \right)^{-0.7} \left( \frac{\theta}{1\arcmin} \right)^2\!\!\!\!,~(\theta > 10\arcsec) \\
\sigma_c & \approx & 2.2 \left( \frac{\nu}{1 \rm~GHz} \right)^{-0.7} \left( \frac{\theta}{1\arcmin} \right)^{10/3}\!\!\!\!\!\!\!\!\!\!,~(\theta < 10\arcsec).
\end{eqnarray}
The source confusion limit is dominated by active galactic nuclei, which have low polarization 
fractions $\lesssim2.5\%$ \citep{Stil2014} and slowly vary with time \citep{Condon1974,Condon2012}.
Therefore both the circularly polarized nature of the emission from hot Jupiters 
and the time variability (discussed in \S\ref{ss:timevariability}) partially mitigate 
the source confusion limit, allowing hot Jupiters, like other time-variable sources,
to be detected below the imaging confusion limit.

The RMS noise ${\sigma}$ in an interferometric observation, 
integration time $\tau$, bandwidth $\Delta \nu$, effective area $A_{\rm eff}$ 
($\approx0.7 \pi (D/2)^2$ for interferometers comprising antennas with dish size $D$),
and $N$ elements is \citep[see e.g.][]{condon+ransom2016} 
\begin{equation}
\sigma = \frac{2 k T_{\rm sys}}{A_{\rm eff} \, \sqrt{\Delta \nu \, \tau N (N-1)}} 
\left(\frac{\rm 10^{26} Jy}{\rm W m^{-2} Hz^{-1}}\right)
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{center}
\begin{deluxetable*}{cccccc}
\tabletypesize{\scriptsize}
\tablecolumns{6}
\tablecaption{Detectability of Hot Jupiters with Current 
and Future Radio Telescopes.\label{tab:sens}}
\tablehead{
Instrument & Band  & Sensitivity\tablenotemark{a} & $\tau$\tablenotemark{b}  & Resolution & $\sigma_c$\tablenotemark{c}\\
           & (MHz) & ($\mu$Jy~hr$^{1/2}$)         &  (hr)                    & (\arcsec)  & ($\mu$Jy bm$^{-1}$)} 
\startdata
GMRT       & 130-190 & 430   &  460             & 16.5 & 56.9 \\ 
HERA       &  30-80  & 870   & 1900             & 1030 & 4.2$\times10^5$\\ 
LOFAR-LBA  &  30-80  & 700   & 1225             & 8.3  & 42 \\ 
LOFAR-HBA  & 110-200 &  35   &    3             & 3.3  & 1.1 \\  
%LWA        &  30-80  & 53000 &  3.2$\times10^6$ & 10.3 & 211 \\
%MWA        &  30-80  & 27000 &  8.1$\times10^5$ & 1030 & $2.1\times10^6$ \\
SKA-Low (part) & 50-200 & 7.8 &  0.15           & 7.8  & 9.7 \\
SKA-Low (full) & 50-350 & 2.2 &  0.012          & 2.2  & 1.5 \\
%VLA 4-Band & 30-80   & 3300  & 1100             & 28.6 & 330 \\
VLA-LOBO   &  50-350 &  92   &  21              & 8.6  & 10
\enddata
\tablecomments{Time estimates to reach 20~$\mu$Jy RMS.}
\tablenotetext{a}{RMS noise in a 1-hour observation, computed for imaging of the full band reported in this table.}
\tablenotetext{b}{The integration time in hours required to reach an RMS of 20~$\mu$Jy (i.e.\ 5-$\sigma$ detection
of a typical source with 100~$\mu$Jy bm$^{-1}$ average flux) \memoTM{Should we compute this instead for a 5-sigma detection
of a $\sim$400~$\mu$Jy bm$^{-1}$ average flux, like in Figure \ref{fig:cartoon}?  Going to 80~$\mu$Jy RMS would be $16\times$ shorter.}.}
\tablenotetext{c}{Source confusion limit RMS contribution.}
\end{deluxetable*}
%\end{center}

% HERA
% http://reionization.org/wp-content/uploads/2015/05/HERA0_imaging.pdf
%
% LOFAR
% https://www.astron.nl/~heald/test/sens.php; 
% 24 core stations, 16 remote, 8 international, 488 subbands
%
% GMRT
% http://gmrt.ncra.tifr.res.in/~astrosupp/obs_setup/sensitivity.html
%

The LOw-Frequency ARray (LOFAR) operates as two separate arrays defined by their 
high and low bands, differing configurations, and antenna designs, which
are called respectively the Low Band Antennas (LBA) and High Band Antennas (HBA).  
LOFAR-LBA operates in the 10-90~MHz range, and LOFAR-HBA operates in the 
100-240~MHz range \citep{vanHaarlem2013}.  We consider it as two separate
instruments here.
LOFAR-LBA's most sensitive band is centered at 60~MHz, and LOFAR-HBA's is at 150~MHz. 
Given the strong frequency dependence of
the effective aperture of a dipole antenna, and that the system noise temperature 
$T_{\rm sys}$ for LOFAR is not fully determined by the sky noise $T_{\rm sky}$, 
we use the LOFAR Image Noise calculator\footnote{
Heald; \url{http://www.astron.nl/~heald/test/sens.php}}, which 
is based on {\it SKA Memo 113}\footnote{
Nijboer, Pandey-Pommier, \& de Bruyn; 
\url{http://www.skatelescope.org/uploaded/59513\_113\_Memo\_Nijboer.pdf}}, for
our fiducial sensitivity and imaging parameters in Table~\ref{tab:sens}.

The Hydrogen Epoch of Reionization Array (HERA\footnote{
\url{http://reionization.org/}}) \ldots

The Giant Metre-wave Radio Telescope (GMRT) is an array comprising thirty 
25-meter antennas, operating at frequencies 130-2000~MHz. The lowest
band (130-190~MHz) exhibits receiver noise $T_{\rm rx}$ comparable to the sky noise.
To be conservative, we use the GMRT online calculator.\footnote{
\url{http://gmrt.ncra.tifr.res.in/~astrosupp/obs\_setup/sensitivity.html}
}

For the low frequency receivers (28-80~MHz; 4-Band) in the % Long Wavelength Array (LWA) and the 
Very Large Array (VLA), the receiver noise temperature $T_{\rm rx}\approx 260~\rm K$ 
\citep{Hicks2012} and is subdominant to $T_{\rm sky}$ by at least -6~dB.  Future
upgrades to the VLA such as the LOw Band Observatory \citep[LOBO][]{Kassim2015IAU},
We estimate the sensitivity of LOBO assuming the receiver noise extrapolates linearly
from $T_{\rm rx}\approx 60~\rm K$ (internal VLA memo) in P-Band (230-470~MHz) to
$T_{\rm rx}\approx 260~\rm K$ in 4-band, and consider just the 50-350~MHz portion
of the full 30-470~MHz band.

We compute $T_{\rm sys} = T_{\rm rx}+ T_{\rm sky}$ using the low frequency sky 
noise temperature fit from \cite{Rogers+Bowman2008}:
\begin{equation}
T_{\rm sky} = T_{150} \left( \frac{\nu}{150~\rm MHz}\right)^{-\beta} + T_{\rm CMB} 
\end{equation}
where $T_{150} = 283.2~\rm K$ is the $T_{\rm sky}$ at 150~MHz, $\beta=2.47$,
and $T_{\rm CMB} = 2.73~\rm K$ is the contribution from the cosmic microwave 
background (CMB).
While this relation was fit to data taken in the 100-200~MHz range, the authors
found it to be consistent with published measurements from 10-408~MHz.

The low frequency component of the  Square Kilometer Array (SKA), recently rebaselined 
to have 0.4~km$^2$ collecting area in 130,000 elements with baselines extending
to 65~km, will operate from 50-350~MHz.\footnote{
\url{https://www.skatelescope.org/news/worlds-largest-radio-telescope-near-construction/}}
We estimate the noise in the lower half of the band ($\nu<200~\rm MHz$), appropriate
for lower mass ($\lesssim 10 M_J$) separately from the full band, which could be used
to detect $\sim 30 M_J$ hot Jupiters.
We adopt the \cite{Rogers+Bowman2008} relation for $T_{\rm sys}$ as above, and assume
$T_{\rm rx}\approx 60~\rm K$.
We caution that these assumptions are optimistic, and the time estimates could in reality 
be a factor of a few worse.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{deluxetable*}{cccccc}
\tabletypesize{\footnotesize}
\tablecolumns{6}
\tablecaption{Detectability of Hot Jupiters with Current 
and Future Radio Telescopes.\label{tab:sens}}
\tablehead{
Instrument & Band  & Sensitivity\tablenotemark{a} & $\tau$\tablenotemark{b}  & Resolution & $\sigma_c$\tablenotemark{c}\\
           & (MHz) & ($\mu$Jy~hr$^{1/2}$)         &  (hr)                    & (\arcsec)  & ($\mu$Jy bm$^{-1}$)} 
\startdata
LOFAR-LBA  &  30-80  & 700   & 120000             & 8.3  & 42 \\ 
LOFAR-HBA  & 110-200 &  35   &    310             & 3.3  & 1.1 \\  
HERA       &  30-80  & 870   & 160000             & 1030 & 4.2$\times10^5$\\ 
GMRT       & 130-190 & 430   &  46000             & 16.5 & 56.9 \\ 
%LWA       &  30-80  & 58000 &  702127141         & 10.3 & 42.2 \\
%MWA       &  30-80  & 29000 &  176223211         & 1030 & 4.2$\times10^5$ \\
SKA-Low (part) & 50-200 & 7.8 &  12.6             & 7.6  & 9.7 \\
SKA-Low (full) & 50-350 & 2.2 &  1.0              & 4.8  & 1.5 \\
%VLA 4-Band& 30-80   & 3300  & 2312058            & 28.6 & 330 \\
VLA-LOBO   &  50-350 &  92   &  1800              & 8.6  & 10.4
\enddata
%\tablecomments{Time estimates to reach 2~$\mu$Jy RMS.}
\tablenotetext{a}{RMS noise in a 1-hour observation, computed for imaging of the full band reported in this table.}
\tablenotetext{b}{The integration time in hours required to reach an RMS of 2~$\mu$Jy (i.e.\ 5-$\sigma$ detection
of a typical source at 100~pc with 10~$\mu$Jy bm$^{-1}$ average flux).}
\tablenotetext{c}{Source confusion RMS contribution.}
\end{deluxetable*}
%
% HERA
% http://reionization.org/wp-content/uploads/2015/05/HERA0_imaging.pdf
%
% LOFAR
% https://www.astron.nl/~heald/test/sens.php; 
% 24 core stations, 16 remote, 8 international, 488 subbands
%
% GMRT
% http://gmrt.ncra.tifr.res.in/~astrosupp/obs_setup/sensitivity.html
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We consider in this section several current and near future radio observatories that 
operate at frequencies $\nu \lesssim 350~\rm MHz$ and can reach continuum 
sensitivities less than 1~mJy in a few hours.  The results are summarized in Table~\ref{tab:sens}.

The primary limitation to detectability is a telescope's sensitivity, characterized by the root mean 
square (RMS) noise fluctuations in the sky and receiver.
The RMS noise ${\sigma}_{\rm RMS}$ in an interferometric observation with 
integration time $\tau$, bandwidth $\Delta \nu$, effective area $A_{\rm eff}$ 
($\approx0.7 \pi (D/2)^2$ for interferometers comprising antennas with dish size $D$),
and $N$ antennas is \citep[see e.g.][]{condon+ransom2016}
%%%
\begin{equation}
\sigma_{\rm RMS} = \frac{2 k T_{\rm sys}}{A_{\rm eff} \, \sqrt{\Delta \nu \, \tau N (N-1)}} 
\left(\frac{\rm 10^{26} Jy}{\rm W m^{-2} Hz^{-1}}\right).
\end{equation}
%%%
Here, $T_{\rm sys}$ is the blackbody temperature equivalent of the total system noise, which is the sum
of the instrument or receiver noise temperature $T_{\rm rx}$ and the noise contribution from the sky $T_{\rm sky}$.
We estimate $T_{\rm sky}$ using the low frequency sky noise temperature fit from \cite{Rogers+Bowman2008}:
%%%
\begin{equation}
T_{\rm sky} = T_{150} \left( \frac{\nu}{150~\rm MHz}\right)^{-\beta} + T_{\rm CMB} 
\end{equation}
%%%
where $T_{150} = 283.2~\rm K$ is the $T_{\rm sky}$ at 150~MHz, $\beta=2.47$,
and $T_{\rm CMB} = 2.73~\rm K$ is the contribution from the cosmic microwave 
background (CMB).
While this relation was fit to data taken in the 100-200~MHz range, the authors
found it to be consistent with published measurements from 10-408~MHz. 
%The $T_{\rm rx}$ is the receiver noise temperature, which depends on the instruments and is described below.

An additional contribution to the noise is due to source confusion, the imaging noise associated with 
unresolved interloping radio sources in an observation.
Source confusion is characterized by the standard deviation $\sigma_c$ in the surface brightness of an 
image due to one or more unresolved sources in the beam solid angle (for a review, see \cite{Condon1974,Condon2012}).
To estimate $\sigma_c$, reported in Table~\ref{tab:sens}, 
we use the relations provided by \cite{condon+ransom2016}, reproduced below:
%%%
\begin{eqnarray}
\sigma_c & \approx & 0.2 \left( \frac{\nu}{1 \rm~GHz} \right)^{-0.7} \left( \frac{\theta}{1\arcmin} \right)^2\!\!\!\!,~(\theta > 10\arcsec) \\
\sigma_c & \approx & 2.2 \left( \frac{\nu}{1 \rm~GHz} \right)^{-0.7} \left( \frac{\theta}{1\arcmin} \right)^{10/3}\!\!\!\!\!\!\!\!\!\!,~(\theta < 10\arcsec).
\end{eqnarray}
%%%
The source confusion limit is dominated by active galactic nuclei, which have low polarization 
fractions $\lesssim2.5\%$ \citep{Stil2014} and slowly vary with time. 
Therefore both the circularly polarized nature of the emission from hot Jupiters 
and the time variability (discussed in \S\ref{ss:timevariability}) partially mitigate 
the source confusion limit, allowing hot jupiters, like other time-variable sources,
to be detected below the imaging confusion limit.  Thus we expect not to be a strong
limitation to detectability of hot jupiters.

% LOFAR
The LOw-Frequency ARray (LOFAR) operates as two separate arrays defined by their 
high and low bands, differing configurations, and antenna designs, which
are called respectively the Low Band Antennas (LBA) and High Band Antennas (HBA).  
LOFAR-LBA operates in the 10-90~MHz range, and LOFAR-HBA operates in the 
100-240~MHz range \citep{vanHaarlem2013}.  We consider it as two separate
instruments here.
LOFAR-LBA's most sensitive band is centered at 60~MHz, and LOFAR-HBA's is at 150~MHz. 
Given the strong frequency dependence of
the effective aperture of a dipole antenna, and that the system noise temperature 
$T_{\rm sys}$ for LOFAR is not fully determined by the sky noise $T_{\rm sky}$ (i.e.\ $T_{\rm rx} \sim T_{\rm sky}$), 
we use the LOFAR Image Noise calculator\footnote{
Heald; \url{http://www.astron.nl/~heald/test/sens.php}}, which 
is based on {\it SKA Memo 113}\footnote{
Nijboer, Pandey-Pommier, \& de Bruyn; 
\url{http://www.skatelescope.org/uploaded/59513\_113\_Memo\_Nijboer.pdf}}, for
our fiducial sensitivity and imaging parameters in Table~\ref{tab:sens}.

% HERA
The Hydrogen Epoch of Reionization Array (HERA\footnote{
\url{http://reionization.org/}}) \ldots

% GMRT
The Giant Metre-wave Radio Telescope (GMRT) is an array comprising thirty 
25-meter antennas, operating at frequencies 130-2000~MHz. The lowest
band (130-190~MHz) exhibits receiver noise $T_{\rm rx}$ comparable to the sky noise.
To be conservative, we use the GMRT online calculator.\footnote{
\url{http://gmrt.ncra.tifr.res.in/~astrosupp/obs\_setup/sensitivity.html}
}

% SKA
The low frequency component of the  Square Kilometer Array (SKA), recently rebaselined 
to have 0.4~km$^2$ collecting area in 130,000 elements with baselines extending
to 65~km, will operate from 50-350~MHz.\footnote{
\url{https://www.skatelescope.org/news/worlds-largest-radio-telescope-near-construction/}}
We estimate the noise in the lower half of the band ($\nu<200~\rm MHz$), appropriate
for lower mass ($\lesssim 10 M_J$) separately from the full band, which could be used
to detect $\sim 30 M_J$ hot Jupiters.
We adopt the \cite{Rogers+Bowman2008} relation for $T_{\rm sys}$ as above, and assume
$T_{\rm rx}\approx 60~\rm K$.
We caution that these assumptions are optimistic, and the time estimates could in reality 
be a factor of a few worse.

% VLA
For the low frequency receivers (28-80~MHz; 4-Band) in the % Long Wavelength Array (LWA) and the 
Very Large Array (VLA), the receiver noise temperature $T_{\rm rx}\approx 260~\rm K$ 
\citep{Hicks2012} and is subdominant to $T_{\rm sky}$ by at least -6~dB.  Future
upgrades to the VLA such as the LOw Band Observatory \citep[LOBO][]{Kassim2015IAU}
are being considered, and would cover the full 30-470~MHz band.  
We estimate the sensitivity of LOBO assuming the receiver noise extrapolates linearly
from $T_{\rm rx}\approx 60~\rm K$ (internal VLA memo) in P-Band (230-470~MHz) to
$T_{\rm rx}\approx 260~\rm K$ in 4-band, and consider just the 50-350~MHz portion
of the full 30-470~MHz band.

The estimates of the noises shown in Table \ref{tab:sens}, together with Figure \ref{fig:observability}, indicate that, at the frequency relevant to the radio emission from Jupiter-size planets, current instrument sensitivities are too low for detection. 
On the other hand, for more massive exoplanets ($M_p\sim 10 M_J$) where the bulk of the radio emission would exist above 50~MHz, there is hope to detect radio emission from nearby systems with LOFAR-HBA or LOBO. 
LOFAR-HBA has reasonable sensitivity to $M_p =10~M_J$ planets at 5 AU around a few closest systems as well as those around further stars but orbiting closer to the stars. 
More promising is the low frequency component of SKA, which will be able to detect the signal within one day from planets with $M_p \sim 10~M_J$ and orbital distances $\sim$5~AU around M stars $<$100~pc from Earth.
